---
title: "RepData1"
author: "Optical Engineer Turning into Data Scientist"
date: "9/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Preprocess the DATA
It is assumed that the zip file of raw data is in the current working directory. After unzipping one file the resulting file will have an *.csv extension. The csv file is read into R as a dataframe called "activity."
The activity dataframe is preprocessed by converting the date variable into a date class.

```{r }
library(dplyr)
library(tidyr)
library(lubridate)
library(xtable)
library(ggplot2)
library(stats)
zipfile <- dir()[grep("zip",dir())]             #get the zip file name
unzip(zipfile)                            #unzip in current directory
activity <- read.csv("activity.csv",sep = ",")
activity$date <- date(activity$date)

```

## What is mean total number of steps taken per day?
### Plot a histogram of the total steps per day

If the lubridate package funcion qday() is applied to the $date variable in the activity dataframe, the day of the quarter is returned. There are 61 unique days. To create the histogram data, the steps for each day are summed. NAs are ignored.
```{r StepInADay}
#add a column for unique date in quarter
activityWQtr <- activity %>% mutate(qtr = qday(date))
dailySums <- aggregate(steps ~ qtr, data = activityWQtr, FUN = sum, na.action = na.omit )
```

Create the histogram:
```{r}
hist(dailySums$steps,
     main = "Frequency of steps per day",
     xlab = "Steps in a day",
     col = rgb(0,1,1,.2))
```

### Mean and median of the total number of steps per day.

```{r}
aveStepsPerDay <- mean(dailySums$steps)
medStepsPerDay <- median(dailySums$steps) 
```

The mean is `r as.numeric(aveStepsPerDay)` steps and the median is `r medStepsPerDay` steps.

