---
title: "RepData1"
author: "GPurvinis"
date: "9/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Preprocess the DATA
It is assumed that the zip file of raw data is in the current working directory. After unzipping one file the resulting file will have an *.csv extension. The csv file is read into R as a dataframe called "activity."
The activity dataframe is preprocessed by converting the date variable into a date class.

```{r }
library(dplyr)
library(tidyr)
library(lubridate)
library(data.table)
library(ggplot2)
library(stats)
zipfile <- dir()[grep("zip",dir())]             #get the zip file name
unzip(zipfile)                            #unzip in current directory
activity <- read.csv("activity.csv",sep = ",")
activity$date <- date(activity$date)

```

## What is mean total number of steps taken per day?
You can also embed plots, for example:

```{r pressure}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
